---
output: github_document
---



```{r echo=FALSE, results="hide", message=FALSE}
library("badger")
```


```{r, echo = FALSE, results='asis'}

cat('# waterlevel',
  '[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)',
  badge_travis("jkennel/waterlevel"),
  '[![Coverage Status](https://img.shields.io/codecov/c/github/jkennel/waterlevel/master.svg)](https://codecov.io/github/jkennel/waterlevel?branch=master)',
	badge_cran_release("waterlevel", "green"),
	badge_cran_download("waterlevel", "grand-total", "green")
)
```

The **waterlevel** package is designed to simplify the analysis of well water level data **R**.  Currently the focus is on barometric/loading efficiency.

# Installation
```{r echo=FALSE, results="hide", message=FALSE}
remotes::install_github("jkennel/waterlevel")
```



## Future work ##

Decrease dependencies
fft methods
Add parameter estimation methods
Add analysis methods (slug testing, pumping tests)
Increase test coverage
Increase memory and computational efficiency
Gap filling methods


