---
output: github_document
always_allow_html: yes
---



```{r echo=FALSE, results="hide", message=FALSE}
library("badger")
```


```{r, echo = FALSE, results='asis'}

cat('# waterlevel',
  '[![DOI](https://zenodo.org/badge/186485037.svg)](https://zenodo.org/badge/latestdoi/186485037)',
  '[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)',
  badge_travis("jkennel/waterlevel"),
  '[![Coverage Status](https://img.shields.io/codecov/c/github/jkennel/waterlevel/master.svg)](https://codecov.io/github/jkennel/waterlevel?branch=master)',
	badge_cran_release("waterlevel", "green"),
	badge_cran_download("waterlevel", "grand-total", "green")
)
```

The **waterlevel** package is designed to simplify the analysis of well water level data in **R**.  Currently the focus is on barometric/loading (BE/LE) efficiency, but the package is under active development.

## Installation

```{r echo=TRUE, eval = FALSE}
remotes::install_github("jkennel/waterlevel")
```

Also, make sure you are using the development version of the *earthtide* package.

```{r echo=TRUE, eval = FALSE}
remotes::install_github("jkennel/earthtide")
```

## Static Barometric/Loading efficiency

```{r echo=TRUE, include = FALSE}
library(waterlevel)
library(data.table)
library(splines)
data(transducer)
```

See the package vignettes for examples the methods.

### Static BE/LE methods 

- be_clark
- be_rahi
- be_least_squares
- be_acworth
- be_high_low
- be_ratio
- be_visual

### Response methods 

- Barometric response functions
    - Regular lags
    - Irregular lags
    - Distributed lags
- Harmonic analysis 
    - Specified frequencies 
    - Specified Earth tide constituents *earthtide* package
- Frequency response function


## Future work ##

- Decrease dependencies
- Other fft methods for frequency response functions
- Add parameter estimation methods
- Add analysis methods (slug testing, pumping tests)
- Increase test coverage
- Increase memory and computational efficiency
- Gap filling methods (interpolation)
- Distributed lag with subset
- factor handling in portion

